---
title: "How to load my model?"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  out.width = "100%"
)

set.seed(42)
torch::torch_manual_seed(42)
```

```{r setup, warning=FALSE, message=FALSE}
library(Survinng)
library(here)
```

The `Survinng` package is designed in a flexible way to allow users to load 
models trained in R or Python using `survivalmodels` or `pycox`, respectively.
Under the hood, the package uses the `torch` package and any base neural network
can be loaded and will be coverted to the corresponding survival model, e.g.,
a DeepHit, DeepSurv, or CoxTime survival model.

## Load a base `torch` model

This is the simplest way to load a model. The model must be a `torch::nn_module`
and any input (tabular, image, multimodal, etc.) can be passed as long as the 
output dimension fits the expected survival deep learning architecture. This
means:

* For a **DeepHit** model, the output dimension must be of shape 
`(batch_size, n_risks, n_time_bins)` where `n_risks` is the number of
competing risks and `n_time_bins` the number of time bins.
* For a **DeepSurv** and **CoxTime** models, the output dimension must be of shape 
`(batch_size, 1)`, since both models are designed to predict the risk of a 
single event and are Cox-based.

### Example DeepHit model

In this example, we will create a simple DeepHit model using the `torch` package.
The model will have 5 competing risks and 20 time bins. The input dimension
will be 10 features. The model will be a simple feedforward neural network with
two hidden layers. Important to note is that the output dimension is reshaped to 
`(batch_size, n_risks, n_time_bins)`.

```{r}
library(torch)

n_risks <- 5
n_time_bins <- 20
n_features <- 10

# Create the base model and data
base_model_deephit <- nn_module(
  initialize = function() {
    self$net <- nn_sequential(
      nn_linear(n_features, 128),
      nn_relu(),
      nn_linear(128, n_risks * n_time_bins)
    )
  },
  forward = function(x) {
    self$net(x) %>%
      torch_reshape(c(-1, n_risks, n_time_bins))
  }
)
model_deephit <- base_model_deephit()
data <- torch_randn(50, n_features)
time_bins <- seq(0, 10, length.out = n_time_bins)

# Check output
output <- model_deephit(data)
print(output$shape)
```

Now, we have our (untrained) base model and corresponding data. The next step
is to create a `Survinng` explainer object. This is done using the `explain()`
function. The function takes the model, data, and time bins (required for
a DeepHit model) as input and internally combines the base model with the
DeepHit architecture. The `model_type` argument is set to `"deephit"` to 
indicate that the model is a DeepHit model. The function will return an 
explainer object which can be used to compute all implemented gradient-based
feature attribution methods in the survival context.

```{r, fig.width= 15, fig.height= 7}
# Create explainer object
exp_deephit <- Survinng::explain(model_deephit, data = data, model_type = "deephit",
                                 time_bins = time_bins)
print(exp_deephit)

# Calculate gradients
grad <- surv_grad(exp_deephit, instance = c(1,3))

# Plot gradients
plot(grad)
```

### Example DeepSurv model

In this example, we will create a simple DeepSurv model using the 
`torch` package. Since DeepSurv is a Cox-based model, the output dimension
will be of shape `(batch_size, 1)` and we need the baseline hazard function.

```{r}
library(torch)

n_features <- 10

# Create baseline hazard function (50 time points)
time <- seq(0, 10, length.out = 50)
hazard <- exp(-(time - 5)^2 / 5)
base_hazard <- data.frame(time = time, hazard = hazard)

# Create the base model and data
base_model_deepsurv <- nn_module(
  initialize = function() {
    self$net <- nn_sequential(
      nn_linear(n_features, 128),
      nn_relu(),
      nn_linear(128, 1, bias = FALSE)
    )
  },
  forward = function(x) {
    self$net(x)
  }
)
model_deepsurv <- base_model_deepsurv()
data <- torch_randn(50, n_features)

# Check output
output <- model_deepsurv(data)
print(output$shape)
```

Now, we have our (untrained) base model and corresponding data, including the
baseline hazard values The next step is to create a `Survinng` explainer
object. This is done using the `explain()` function. The function takes the
model, data, and baseline hazard (required for a DeepSurv model) as input
and internally combines the base model with the DeepSurv architecture. The
`model_type` argument is set to `"deepsurv"` to indicate that the model is a
DeepSurv model. 

```{r, fig.width= 15, fig.height= 7}
# Create explainer object
exp_deepsurv <- Survinng::explain(model_deepsurv, data = data, 
                                  model_type = "deepsurv",
                                  baseline_hazard = base_hazard)
print(exp_deepsurv)

# Calculate IntGrad(t)
ig <- surv_intgrad(exp_deepsurv, instance = c(1,3))

# Show results as a contribution plot
plot(ig, type = "contr")
```


### Example CoxTime model

In this example, we will create a simple CoxTime model using the `torch` package.
It is very similar to the DeepSurv modelm i.e., the output dimension of the 
base model will be of shape `(batch_size, 1)` and we need the baseline hazard 
function since it is a Cox-based model. However, the time points are integrated 
to the input features, so the model will have `n_features + 1` input features. 
Currently, the default function for adding the time points to the input features
is the following one and assumes tabular data:

```{r}
n_features <- 10
x <- torch_randn(50, n_features)
t <- torch_linspace(0, 10, steps = 50)

preprocess_fun <- function(x) {
  batch_size <- x$size(1)
  
  # Input (batch_size, n_features) -> (batch_size * t, n_features) (replicate each row t times)
  res <- x$repeat_interleave(length(t), dim = 1)
  
  # Add time to input
  time <- torch_vstack(replicate(batch_size, t$unsqueeze(-1)))
  
  # Combine both and return
  torch::torch_cat(list(res, time), dim = -1)
}

# Input shape
print(x$shape)

# Shape after preprocessing
preprocess_fun(x)$shape
```

However, you can pass your own preprocessing function to the `explain()` function
using the `preprocess_fun` argument, but this option is still experimental and
not fully tested. 

In the following code, we create the base model and data for the CoxTime
architecture. The model will have 10 + 1 features and 50 time points for the baseline
hazard function (this is basically the same as the DeepSurv model from 
[this Section](#example-deepsurv-model)):

```{r}
library(torch)
n_features <- 10

# Create baseline hazard function (50 time points)
time <- seq(0, 10, length.out = 50)
hazard <- exp(-(time - 5)^2 / 5)
base_hazard <- data.frame(time = time, hazard = hazard)

# Create the base model and data
base_model_coxtime <- nn_module(
  initialize = function() {
    self$net <- nn_sequential(
      nn_linear(n_features + 1, 128),
      nn_relu(),
      nn_linear(128, 1, bias = FALSE)
    )
  },
  forward = function(x) {
    self$net(x)
  }
)
model_coxtime <- base_model_coxtime()
data <- torch_randn(50, n_features)

# Check output
output <- model_coxtime(preprocess_fun(data))
print(output$shape) # the instances are replicated t times
```

Now, we have our (untrained) base model and corresponding data, including the
baseline hazard values. The next step is to create a `Survinng` explainer
object. This is done using the `explain()` function. The function takes the
model, data, and baseline hazard (required for a CoxTime model) as input
and internally combines the base model with the CoxTime architecture. The
`model_type` argument is set to `"coxtime"` to indicate that the model is a
CoxTime model. The `preprocess_fun` argument is set to the function we defined
earlier to add the time points to the input features. 

```{r, fig.width= 15, fig.height= 7}
# Create explainer object
exp_coxtime <- Survinng::explain(model_coxtime, data = data, 
                                 model_type = "coxtime",
                                 baseline_hazard = base_hazard)
print(exp_coxtime)

# Calculate IntGrad(t)
ig <- surv_intgrad(exp_coxtime, instance = c(1,3))

# Show results as a contribution plot
plot(ig, type = "contr")
```


## Load a `survivalmodels` model

The `survivalmodels` package is a R package for survival analysis using deep
learning. Especially for the deep learning models, it uses the `pycox` package 
in Python under the hood via `reticulate`. The `Survinng` package is designed
to load models trained in `survivalmodels` directly. However, there are some
issues when `torch` in R (i.e., `LibTorch`) and `Pytorch` in Python are 
loaded in the same session (see [issue #815](https://github.com/mlverse/torch/issues/815)).
Therefore, we keep the training procedure using `survivalmodels` in a separate
session using `callr`. 

In the following example, we will train a default DeepHit, DeepSurv and Coxtime
model using the `survivalmodels` package. The models will be trained on the
`lung` dataset from the `survival` package. 

```{r}
library(survival)
library(callr)

# Load lung dataset
data(cancer, package = "survival")
data <- na.omit(cancer[, c(1, 4, 5, 6, 7, 10, 2, 3)])

# Train models in separate session
models <- callr::r(function(data) {
  library(survivalmodels)
  library(survival)
  
  # Set seed for reproducibility
  set.seed(42)
  set_seed(42)
  
  # Make sure pycox is installed
  install_pycox(install_torch = TRUE)
  
  # Fit the DeepSurv model
  fit_deepsurv <- deepsurv(Surv(time, status) ~ ., data = data, epochs = 10,
                           num_nodes = 8L)
  
  # Fit the DeepHit model
  fit_deephit <- deephit(Surv(time, status) ~ ., data = data, epochs = 10,
                         cuts = 30, num_nodes = 8L)
  
  # Fit the CoxTime model
  fit_coxtime <- coxtime(Surv(time, status) ~ ., data = data, epochs = 10,
                         num_nodes = 8L)

  # Extract the models using `Survinng`
  ext_deepsurv <- Survinng::extract_model(fit_deepsurv)
  ext_deephit <- Survinng::extract_model(fit_deephit)
  ext_coxtime <- Survinng::extract_model(fit_coxtime)
  
  # Return the extracted models
  list(deepsurv = ext_deepsurv, deephit = ext_deephit, coxtime = ext_coxtime)
}, args = list(data = data))
```


Now, we have our trained models in the `models` list. The models are
extracted using the `extract_model()` function from the `Survinng` package.
These objects and the data to be explained can be passed to the `explain()` 
function to create the `Survinng` explainer objects. 

```{r}
# Create explainer objects
exp_deepsurv <- Survinng::explain(models$deepsurv, data = data)
exp_deephit <- Survinng::explain(models$deephit, data = data)
exp_coxtime <- Survinng::explain(models$coxtime, data = data)
```

The `explain()` function will automatically detect the model type and
create the corresponding explainer object, i.e., the `model_type` argument is
not needed in this case. Using the explainer objects, we can compute the
gradient-based feature attribution methods. The `instance` argument is used
to specify the instances for which the explanations should be computed.

```{r, fig.width= 15, fig.height = 7}
# Show instances to be explained
ids <- c(1, 76)
data[ids, ]

# Calculate feature attribution methods
grad <- surv_grad(exp_deepsurv, instance = ids)
ig <- surv_intgrad(exp_deephit, instance = ids)
shap <- surv_gradSHAP(exp_coxtime, instance = ids)

# Show the predictions
plot(shap, type = "pred")
```


With the `plot()` function, we can visualize the results. The `type` argument
can be set to `"pred"` to show the predictions of the model, `"attr"` to show 
the feature attribution values, or `"contr"` to show the contribution plots.

```{r, fig.width= 12, fig.height = 7}
library(cowplot)
library(ggplot2)

plot_grid(
  plot(ig, type = "contr", aggregate = TRUE) + ggtitle("IntGrad(t) (aggregated)"),
  plot(shap, type = "attr") + ggtitle("SHAP"),
  ncol = 2
)
```


## Load a `pycox` Python model

It is also possible to load models trained in Python using the `pycox` package.
Basically, it is the same procedure as for another 
[`torch::nn_module` model](#load-a-base-torch-model). The only difference is 
that the model needs to be loaded in R as described in the [Pyton to R article](https://torch.mlverse.org/docs/articles/python-to-r) for the `torch` package.
This means, the model architecture must be rebuilt in R and the weights and 
parameters must be loaded from the Python model.

For illustration, we will use the same CoxTime model from the `pycox` Jupyter notebook
[here](https://github.com/havakv/pycox/blob/master/examples/cox-time.ipynb) on
the METABRIC dataset. After training the model, follow these steps to extract
the base hazard function and the model weights (`model` and `x_test` are defined
as in the mentioned notebook): 

```{python, eval = FALSE}
import pandas as pd

# We use only a sample of the data for the baseline hazard
base_hazard = model.compute_baseline_hazards(sample = 200)

# Save base hazard as csv
pd_base_hazard = pd.DataFrame({
    'time': base_hazard.index.to_numpy(), 
    'hazard': base_hazard.to_numpy()})
pd_base_hazard.to_csv('base_hazard.csv', index = False)

# Save model weights (not the  whole model!)
torch.save(model.net.net.state_dict(), 'model.pt')

# Save data to be explained
df_train.to_csv('df_train.csv', index=False)
pd.DataFrame(x_val, columns = df_val.columns[0:9]).to_csv('x_val.csv', index=False)
```

Now, the tricky part is rebuilding the model architecture in R with the same
names and parameters as in the Python model. This is done in the following code:

```{r}
# Rebuild a dense layer
BaseMLP_dense <- nn_module(
  initialize = function(in_feat, out_feat) {
    self$linear <- nn_linear(in_feat, out_feat)
    self$batch_norm <- nn_batch_norm1d(out_feat)
  },

  forward = function(x) {
    x <- self$linear(x)
    x <- nn_relu()(x)
    x <- self$batch_norm(x)
  }
)

# Rebuild the whole model
BaseMLP <- torch::nn_module(
  classname = "BaseMLP",
  initialize = function() {
    
    self$net <- nn_sequential(
      BaseMLP_dense(10, 32),
      BaseMLP_dense(32, 32),
      nn_linear(32, 1, bias = FALSE)
    )
  },
  forward = function(x) {
    if (is.list(x)) x <- x[[1]]
    self$net(x)
  }
)

# Create model and load state dict
model <- BaseMLP()
state_dict <- load_state_dict(here("vignettes/articles/how_to_load_model/model.pt"))
model$load_state_dict(state_dict)
model$eval()
print(model)
```

The model architecture is now rebuilt in R and the weights are loaded from
the Python model. The next step is to load the baseline hazard function and
the data to be explained. Additionally, since in the notebook uses a label
transformation, we need to rebuild the transformation function in R. However,
this is a simple mean and standard deviation transformation, so we can use the
`labtrans` argument in the `explain()` function to define the transformation
function. The `labtrans` argument is a list with two functions: `transform`
and `transform_inv`.

```{r}
# Load base hazard and data
base_hazard <- read.csv(here("vignettes/articles/how_to_load_model/base_hazard.csv"))
df_train <- read.csv(here("vignettes/articles/how_to_load_model/df_train.csv"))
x_val <- read.csv(here("vignettes/articles/how_to_load_model/x_val.csv"))

# Rebuild the label transformation
mean <- mean(df_train$duration)
sd <- sd(df_train$duration)
labtrans <- list(
  transform = function(x) (x - mean) / sd,
  transform_inv = function(x) x * sd + mean
)
```

Now, we can create the `Survinng` explainer object using the `explain()` function
and apply all implemented gradient-based feature attribution methods on this 
model. The `model_type` argument is set to `"coxtime"` to indicate that the
model is a CoxTime model and the `labtrans` argument is set to the defined 
transformation.

```{r, fig.width= 10, fig.height= 15}
# Create explainer
explainer <- Survinng::explain(model, data = x_val, model_type = "coxtime",
                               baseline_hazard = base_hazard,
                               labtrans = labtrans)

# Calculate GradSHAP(t)
grad_shap <- surv_gradSHAP(explainer, instance = c(1, 3))

# Show predictions
p1 <- plot(grad_shap, type = "pred")

# Show attribution results
p2 <- plot(grad_shap, type = "attr", add_comp = c("pred_diff"))

# Show contribution percentages
p3 <- plot(grad_shap, type = "contr")

cowplot::plot_grid(p1, p2, p3, ncol = 1)
```

In a similar way, you can also load DeepSurv and DeepHit models from
`pycox`. The only difference is that the model architecture must be
changed accordingly. The `model_type` argument is set to `"deepsurv"` or
`"deephit"` to indicate the model type.
