---
title: "Methods for Feature Attributions in Survival Models"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```

```{r setup}
library(Survinng)
library(ggplot2)
set.seed(42)
```

# Time-independent effects

## Generate the data

Simulation setting:
- $10,000$ samples ($9,500$ for training, $500$ for testing)
- No time-dependent effects
- $X_1 \sim \mathcal{N}(0,1)$ has a small positive effect on the hazard -> negative effect on survival
- $X_2 \sim \mathcal{U}(0,1)$ has a strong negative effect on the hazard -> positive effect on survival
- $X_3 \sim \mathcal{U}(-1,1)$ has no effect

```{r sim1 data generation}
library(survival)
library(simsurv)

# Simulate data
n <- 10000
x <- data.frame(x1 = rnorm(n), x2 = runif(n, 0, 1), x3 = runif(n, -1, 1))
simdat <- simsurv(dist = "weibull", lambdas = 0.1, gammas = 2.5, betas = c(x1 = 1, x2 = -2),
                  x = x, maxt = 10)
y <- simdat[, -1]
colnames(y)[1] <- "time"
dat <- cbind(y, x)

# Train/test
idx <- sample(n, 9500)
train <- dat[idx, ]
test <- dat[-idx, ]
```

## Fit the models
```{r utils1, echo = FALSE}
fit_model <- function(model, train, test) {
  callr::r(function(model, train, test) {
    library(survivalmodels)
    library(survival)
    reticulate::use_condaenv("Survinng", required = TRUE)
    set.seed(1)
    set_seed(1)
    
    # Fit model
    if (model == "CoxTime") {
      model <- coxtime(Surv(time, status) ~ ., data = train, verbose = FALSE, epochs = 500L,
                     early_stopping = TRUE, frac = 0.33, batch_size = 1024L, patience = 10L,
                     dropout = 0.1)
    } else if (model == "DeepHit") {
      model <- deephit(Surv(time, status) ~ ., data = train, verbose = FALSE, epochs = 500L,
                     early_stopping = TRUE, frac = 0.33, patience = 10L, cuts = 30, 
                     batch_size = 1024L, dropout = 0.1)
    } else if (model == "DeepSurv") {
      model <- deepsurv(Surv(time, status) ~ ., data = train, verbose = FALSE, epochs = 100L,
                      early_stopping = TRUE, frac = 0.33, patience = 10L, batch_size = 1024L,
                      dropout = 0.1)
    } else {
      stop("Model not found")
    }
    
    # Make predictions
    pred <- predict(model, newdata = test, type = "survival")
    dat <- data.frame(
      pred = c(pred), 
      time = rep(as.numeric(colnames(pred)), each = nrow(pred)),
      id = rep(1:nrow(pred), ncol(pred))
    )
    list(Survinng::extract_model(model), pred = dat)
  }, list(model, train, test))
}
```

```{r sim1 fit models, fig.width=10, fig.height=15}
# Fit the models
ext_deephit <- fit_model("DeepHit", train, test)
ext_coxtime <- fit_model("CoxTime", train, test)
ext_deepsurv <- fit_model("DeepSurv", train, test)
```


## Create Explainer

```{r sim1 explainer}
library(Survinng)
library(torch)

# Create explainer
exp_deephit <- explain(ext_deephit[[1]], data = test)
exp_coxtime <- explain(ext_coxtime[[1]], data = test)
exp_deepsurv <- explain(ext_deepsurv[[1]], data = test)
```


## Survival Attribution
```{r sim1 patients}
head(test)
```

### Gradient x Input


```{r sim1 gradient, fig.width=10, fig.height=15}
library(ggplot2)
library(data.table)

df <- melt(data.table(train), id.vars = c("time", "status"))
df_inst <- melt(data.table(test, id = paste0(rownames(train), " (row: " , seq_len(nrow(train)), ")"))[c(1,2), ], 
                id.vars = c("time", "status", "id"))
ggplot(data = df, aes(x = value)) + 
  geom_point(data = df_inst, mapping = aes(x = value, y = 0, color = id)) +
  geom_density(alpha = 0.5, fill = "gray75") +
  theme_minimal() +
  facet_grid(cols = vars(variable), scales = "free")

grad_cox <- surv_grad(exp_coxtime, instance = c(1,2), times_input = TRUE)
grad_deephit <- surv_grad(exp_deephit, instance = c(1,2), times_input = TRUE)
grad_deepsurv <- surv_grad(exp_deepsurv, instance = c(1,2), times_input = TRUE)

cowplot::plot_grid(
  plot(grad_cox), 
  plot(grad_deephit), 
  plot(grad_deepsurv),
  nrow = 3, labels = c("CoxTime", "DeepHit", "DeepSurv"))
```

**Explanation:**

* Gradient x Input uses a reference value of $0$, implicitly. However, we can
assume that $f(t, 0) \approx 0$ holds.
* Instance 1:
  * $X_1 = -0.8$ and $\beta_1 = 1$ (i.e., $X_1 \beta_1 < 0$) -> positive 
  contribution on the survival function (compared to the zero baseline)
  * $X_2 = 0.8$ and $\beta_2 = -2$ (i.e., $X_2 \beta_2 <<0$) -> strong positive
  contribution on the survival function
  * $X_3 = -0.7$ and $\beta_3 = 0$ (i.e., $X_3 \beta_3 = 0$) -> no contribution
* Instance 2:
  * $X_1 = 0.25$ and $\beta_1 = 1$ (i.e., $X_1 \beta_1 > 0$) -> negative 
  contribution on the survival function (compared to the zero baseline)
  * $X_2 = 0.97$ and $\beta_2 = -2$ (i.e., $X_2 \beta_2 <<0$) -> strong positive
  contribution
  * $X_3 = 0.5$ and $\beta_3 = 0$ (i.e., $X_3 \beta_3 = 0$) -> no contribution


### SmoothGrad x Input

```{r sim1 smoothgrad, fig.width=10, fig.height=15}
sg_cox <- surv_smoothgrad(exp_coxtime, instance = c(1,2), n = 50, noise_level = 0.3,
                          times_input = TRUE)
sg_deephit <- surv_smoothgrad(exp_deephit, instance = c(1,2), n = 50, noise_level = 0.3,
                              times_input = TRUE)
sg_deepsurv <- surv_smoothgrad(exp_deepsurv, instance = c(1,2), n = 50, noise_level = 0.3,
                               times_input = TRUE)

cowplot::plot_grid(
  plot(sg_cox), 
  plot(sg_deephit), 
  plot(sg_deepsurv),
  nrow = 3, labels = c("CoxTime", "DeepHit", "DeepSurv"))
```

### IntegratedGradient

#### Zero baseline (should be proportional to Gradient x Input)

```{r sim1 smoothgrad, fig.width=10, fig.height=15}
x_ref <- matrix(c(0,0,0), nrow = 1)
ig_cox <- surv_intgrad(exp_coxtime, instance = c(1,2), n = 50, x_ref = x_ref)
ig_deephit <- surv_intgrad(exp_deephit, instance = c(1,2), n = 50, x_ref = x_ref)
ig_deepsurv <- surv_intgrad(exp_deepsurv, instance = c(1,2), n = 50, x_ref = x_ref)

cowplot::plot_grid(
  plot(ig_cox), 
  plot(ig_deephit), 
  plot(ig_deepsurv),
  nrow = 3, labels = c("CoxTime", "DeepHit", "DeepSurv"))
```

#### Mean baseline

```{r sim1 smoothgrad, fig.width=10, fig.height=15}
x_ref <- NULL # default: feature-wise mean
ig_cox <- surv_intgrad(exp_coxtime, instance = c(1,2), n = 50, x_ref = x_ref)
ig_deephit <- surv_intgrad(exp_deephit, instance = c(1,2), n = 50, x_ref = x_ref)
ig_deepsurv <- surv_intgrad(exp_deepsurv, instance = c(1,2), n = 50, x_ref = x_ref)

cowplot::plot_grid(
  plot(ig_cox), 
  plot(ig_deephit), 
  plot(ig_deepsurv),
  nrow = 3, labels = c("CoxTime", "DeepHit", "DeepSurv"))
```


# Time-dependent effects

## Generate the data

Simulation setting:
- $10,000$ samples ($9,500$ for training, $500$ for testing)
- $X_1 \sim \mathcal{N}(0,1)$ has a small positive effect on the hazard -> negative effect on survival
- $X_2 \sim \mathcal{U}(0,1)$ has a time-dependent effect: first negative on hazard 
and after $t = 2$ positive -> first positive one survival and after $t = 2$ negative
- $X_3 \sim \mathcal{U}(-1,1)$ has no effect

```{r sim2 data generation}
set.seed(42)
library(survival)
library(simsurv)

# Simulate data
n <- 10000
x <- data.frame(x1 = rnorm(n), x2 = runif(n, 0, 1), x3 = runif(n, -1, 1))
simdat <- simsurv(dist = "weibull", lambdas = 0.01, gammas = 4, betas = c(x1 = 0.25, x2 = -1),
                  x = x, maxt = 8, tde = c(x2 = 2), tdefunction = function(x) exp(x - 2) / (1 + exp(x - 2)))
y <- simdat[, -1]
colnames(y)[1] <- "time"
dat <- cbind(y, x)

# Train/test
idx <- sample(n, 9500)
train <- dat[idx, ]
test <- dat[-idx, ]
```

## Fit the models

```{r sim1 fit models, fig.width=10, fig.height=15}
# Fit the models
ext_deephit <- fit_model("DeepHit", train, test)
ext_coxtime <- fit_model("CoxTime", train, test)
ext_deepsurv <- fit_model("DeepSurv", train, test)
```


## Create Explainer

```{r sim1 explainer}
library(Survinng)
library(torch)

# Create explainer
exp_deephit <- explain(ext_deephit[[1]], data = test)
exp_coxtime <- explain(ext_coxtime[[1]], data = test)
exp_deepsurv <- explain(ext_deepsurv[[1]], data = test)
```


## Survival Attribution
```{r sim1 patients}
head(test)
```

### Gradient x Input

```{r sim1 gradient, fig.width=10, fig.height=15}
grad_cox <- surv_grad(exp_coxtime, instance = c(1, 19), times_input = TRUE)
grad_deephit <- surv_grad(exp_deephit, instance = c(1, 19), times_input = TRUE)
grad_deepsurv <- surv_grad(exp_deepsurv, instance = c(1,19), times_input = TRUE)

cowplot::plot_grid(
  plot(grad_cox) + ylim(-0.15, 0.25), 
  plot(grad_deephit) + ylim(-0.15, 0.25), 
  plot(grad_deepsurv) + ylim(-0.15, 0.25),
  nrow = 3, labels = c("CoxTime", "DeepHit", "DeepSurv"))
```


### IntegratedGradient

#### Zero baseline (should be proportional to Gradient x Input)

```{r sim1 smoothgrad, fig.width=10, fig.height=15}
x_ref <- matrix(c(0,0,0), nrow = 1)
ig_cox <- surv_intgrad(exp_coxtime, instance = c(1,19), n = 50, x_ref = x_ref)
ig_deephit <- surv_intgrad(exp_deephit, instance = c(1,19), n = 50, x_ref = x_ref)
ig_deepsurv <- surv_intgrad(exp_deepsurv, instance = c(1,19), n = 50, x_ref = x_ref)

cowplot::plot_grid(
  plot(ig_cox), 
  plot(ig_deephit), 
  plot(ig_deepsurv),
  nrow = 3, labels = c("CoxTime", "DeepHit", "DeepSurv"))
```

#### Mean baseline

```{r sim1 smoothgrad, fig.width=10, fig.height=15}
x_ref <- NULL # default: feature-wise mean
ig_cox <- surv_intgrad(exp_coxtime, instance = c(1,19), n = 50, x_ref = x_ref)
ig_deephit <- surv_intgrad(exp_deephit, instance = c(1,19), n = 50, x_ref = x_ref)
ig_deepsurv <- surv_intgrad(exp_deepsurv, instance = c(1,19), n = 50, x_ref = x_ref)

cowplot::plot_grid(
  plot(ig_cox), 
  plot(ig_deephit), 
  plot(ig_deepsurv),
  nrow = 3, labels = c("CoxTime", "DeepHit", "DeepSurv"))
```

### GradShap

```{r sim1 smoothgrad, fig.width=10, fig.height=15}
gshap_cox <- surv_gradSHAP(exp_coxtime, instance = c(1,19), n = 50, num_samples = 100, 
                           batch_size = 1000)
gshap_deephit <- surv_gradSHAP(exp_deephit, instance = c(1,19), n = 50, num_samples = 100, 
                           batch_size = 1000)
gshap_deepsurv <- surv_gradSHAP(exp_deepsurv, instance = c(1,19), n = 50, num_samples = 100, 
                           batch_size = 1000)

cowplot::plot_grid(
  plot(gshap_cox), 
  plot(gshap_deephit), 
  plot(gshap_deepsurv),
  nrow = 3, labels = c("CoxTime", "DeepHit", "DeepSurv"))
```


### What about interactions?

```{r sim1 smoothgrad, fig.width=10, fig.height=15}
inthess_coxtime <- surv_inthess(exp_coxtime, instance = c(1,19), n = 20, batch_size = 1000)
inthess_deephit <- surv_inthess(exp_deephit, instance = c(1,19), n = 20, batch_size = 1000)
inthess_deepsurv <- surv_inthess(exp_deepsurv, instance = c(1,19), n = 20, batch_size = 1000)

cowplot::plot_grid(
  plot(inthess_coxtime, add_sum = TRUE), 
  plot(inthess_deephit, add_sum = TRUE), 
  plot(inthess_deepsurv, add_sum = TRUE),
  nrow = 3, labels = c("CoxTime", "DeepHit", "DeepSurv"))
```
